---
title: "Theme \"corrobs\""
author: "M. Denis"
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
colorlinks: true
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: true
  word_document: default
---

<!--
This R chunk is used to set up important options and load required packages.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")#, fig.width=12, fig.height=8, fig.path=figures/")
```


# Statistical modelling

## References

* [Pinheiro and Bates (2006)](http://www.worldcat.org/oclc/54860997)

* [Ruiz-Cardenaz et al. (2012)](http://dx.doi.org/10.1016/j.csda.2011.10.024)

* [Blangiardo and Cameletti (2015)](http://www.worldcat.org/oclc/900278105)


## Notations

* $N$: number of individuals

* $T$: number of time points (same for all individuals)

* $y_{it}$: measurement on individual $i$ at time $t$

* $\mu_i$: intercept for individual $i$

* $\rho$: correlation between measurements at two consecutive time points

* $\epsilon_{it}$: measurement error on individual $i$ at time $t$

* $\sigma^2$: variance of the errors


## Vector auto-regression of order 1 (VAR1)

* data: $\mathcal{D} = \{y_{it}\}_{i=1,\ldots,N;t=1,\ldots,T}$

* parameters: $\Theta = \{\mu_1, \ldots, \mu_N, \rho, \sigma^2\}$

* likelihood: $\mathcal{L}(\Theta) = f(\mathcal{D} | \Theta)$

\begin{align}
\forall i, t > 1, \; \; y_{it} &= \rho \, y_{i(t-1)} +\epsilon_{it} \text{ with } \epsilon_{it} \sim \mathcal{N}(0, \sigma^2) \\
\forall i, \; \; y_{i1} &\sim \mathcal{N} \left(0, \frac{\sigma^2}{(1-\rho^2)} \right) \text{ with } |\rho| < 1 \\
\end{align}

<!--
In matrix form:

\[
Y = U Z + E \text{ with } E \sim \mathcal{N}_{N \times (T-1)}(0, \sigma^2 I, 0)
\]

where $Y$ is $N \times (T-1)$, $U$ is $N \times (N+1)$, $Z$ is $(N+1) \times (T-1)$ and $E$ is $N \times (T-1)$, such that:

\[
Y = \begin{bmatrix}
y_{12} & \cdots & y_{1T} \\
\vdots & \ddots & \vdots \\
y_{N2} & \cdots & y_{NT} \\
\end{bmatrix},
\; \; U = \begin{bmatrix}
\mu_1 & \rho & 0 & \cdots & 0 \\
\vdots & 0 & \rho &  & \vdots \\
\vdots & \vdots &  & \ddots &  \\
\mu_N & 0 & \cdots &  & \rho \\
\end{bmatrix},
\; \; Z = \begin{bmatrix}
1 & \cdots & 1 \\
y_{11} & \cdots & y_{1(T-1)} \\
\vdots & \ddots & \vdots \\
y_{N1} & \cdots & y_{N(T-1)} \\
\end{bmatrix}.
\]
-->


## First order univariate dynamic linear model (DLM1)

* data: $\mathcal{D} = \{y_{t}\}_{t=1,\ldots,T}$

* parameters: $\Theta = \{\rho, \sigma_x^2, \sigma_y^2\}$

* latent field: $\xi = \{x_{t}\}_{t=1,\ldots,T}$

* likelihood: $\mathcal{L}(\Theta) = f(\mathcal{D} | \Theta)$

\begin{equation}
\left\{
\begin{array}{lcc}
y_t = x_t +v_t & v_t \sim N(0, \sigma_y^2)& t=1,\, \dots, n\\ 
x_t = \rho x_{t-1} +w_t &  w_t \sim N(0, \sigma_x^2)& t=2,\, \dots, n \\
\end{array}
\right.
\end{equation}


## Linear mixed model with autocorrelated errors (LMMcorr)

* data: $\mathcal{D} = \{y_{it}\}_{i=1,\ldots,N;t=1,\ldots,T}$
  
* parameters: $\Theta = \{ \mu, \rho, \sigma_{\varepsilon}^2, \sigma_a^2\}$
    
* likelihood: $\mathcal{L}(\Theta) = f(\mathcal{D} | \Theta)$

\begin{align}
\forall i, t, \; y_{it} &= \mu + a_i + \varepsilon_{it} \\
\forall i, \; a_i &\sim \mathcal{N}(0, \sigma_a^2) \\
\forall i, \; \varepsilon_{i1} &\sim \mathcal{N}(0, \sigma_{\varepsilon}^2) \\
\forall i, t > 1, \; \varepsilon_{it} &= \rho \, \varepsilon_{i(t-1)} +s_{it} \text{ with }	s_{it} \sim \mathcal{N}(0, \sigma^2_{\varepsilon}(1-\rho^2)) \text{ and } |\rho| < 1 \\
\end{align}


# Simulation

## Help

The command-line `model4all_simul --theme corrobs --themhlp` returns:

```{r, echo=FALSE}
cat(system(command="../src/model4all_simul --theme corrobs --themhlp", intern=TRUE), sep="\n")
```


## Case studies

Simulate phenotypes according to the vector auto-regression of order 1:

```{r corrobs_simul_VAR1, eval=FALSE}
model="VAR1"; model4all_simul --theme corrobs --task corrobs-${model} \
--simd corrobs_simul --args "--model $model"
```

Simulate phenotypes according to the first order univariate dynamic linear model:

```{r corrobs_simul_DLM1, eval=FALSE}
model="DLM1"; model4all_simul --theme corrobs --task corrobs-${model} \
--simd corrobs_simul --args "--model $model"
```

Simulate phenotypes according to the linear mixed model with autocorrelated errors:

```{r corrobs_simul_LMMcorr, eval=FALSE}
model="LMMcorr"; model4all_simul --theme corrobs --task corrobs-${model} \
--simd corrobs_simul --args "--model $model"
```


## TODO

* simulations from Poisson and Binomial distribution

* make $\rho$ depends on spatial structure



# Inference

## Help

The command-line `model4all_infer --theme corrobs --themhlp` returns:

```{r, echo=FALSE}
cat(system(command="../src/model4all_infer --theme corrobs --themhlp", intern=TRUE), sep="\n")
```


## Case studies

Perform inference for the vector auto-regression of order 1, for example with the nlme package:

```{r corrobs_infer_nlme, eval=FALSE}
pkg="nlme"; model="VAR1"; model4all_infer --theme corrobs --task corrobs-${model} \
--simd corrobs_simul --infd corrobs_infer_$pkg --args "--pkg $pkg"
```

Perform inference for the first order univariate dynamic linear model, for example with the INLA package:

```{r corrobs_infer_INLA, eval=FALSE}
pkg="INLA"; model="DLM1"; model4all_infer --theme corrobs --task corrobs-${model} \
--simd corrobs_simul --infd corrobs_infer_$pkg --args "--pkg $pkg --model $model"
```


# Evaluation

...



# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
