---
title: "Introduction to \"model4all\""
author: "T. Flutre & M. Denis"
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
colorlinks: true
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: true
  word_document: default
---

<!--
This R chunk is used to set up important options and load required packages.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")#, fig.width=12, fig.height=8, fig.path=figures/")
```


# Preamble

This document was generated from a [plain-text](https://en.wikipedia.org/wiki/Plain_text) file in the [R Markdown](http://rmarkdown.rstudio.com/) format (Rmd), especially useful with the [R](https://en.wikipedia.org/wiki/R_%28programming_language%29) software (use a recent-enough version, e.g. 3.0, with the [rmarkdown](https://cran.r-project.org/web/packages/rmarkdown/) package).

The Rmd format allows to easily combine different components: regular paragraphs and section headings, but also chunks of code, as well as equations written in [LaTeX](https://en.wikipedia.org/wiki/LaTeX).
A typical Rmd file can then be "converted" into a PDF or HTML file via the `rmarkdown` package which uses [pandoc](https://en.wikipedia.org/wiki/Pandoc) (a recent enough version of pandoc is recommended, such as 1.17).
Such documents hence are called "dynamic reports": the PDF or HTML file can be regenerated whenever the underlying Rmd file changes.

Practically, any Rmd file is best edited with [RStudio](https://en.wikipedia.org/wiki/RStudio), or [Emacs](https://en.wikipedia.org/wiki/Emacs) (with the following packages: [ESS](https://en.wikipedia.org/wiki/Emacs_Speaks_Statistics), [markdown](http://jblevins.org/projects/markdown-mode/) and [polymode](https://github.com/vitoshka/polymode)).



# Overview

This document is part of the "model4all" project.
In the search for statistical models better adjusted to large, heterogeneous data sets, it is always fruitful to "experiment", that is, to simulate some not-too-unrealistic data sets, to fit various models, to compare them in terms of inference and prediction, to assess multiple software tools, etc.

The "model4all" project aims at providing scientists, notably in plant biology, who are willing to learn more about statistical modelling, a light infrastructure facilitating such comparisons.
By using [free and open-source software](https://en.wikipedia.org/wiki/Free_and_open-source_software) and being available on [GitHub](https://github.com/timflutre/model4all), it also aims at promoting collaborative development of such comparisons between scientists.



# Functioning

A **theme** gathers a set of simulation and inferential models which are used to explore a common research question.
For instance, the "quantgen" theme is concerned with quantitative genetics, the "corrobs" theme deals with correlated observations, etc.

For any given theme, the framework is decomposed into 3 steps:

1. **simulate** some data according to a given model;

2. **inference** by fitting a given model (possibly different than at step 1) to these data;

3. **evaluate** parameter estimation, model fit, prediction, running time, memory usage, etc.

Each step has a corresponding **generic program** located in the `src` directory of the project.
After having followed the "Installation" instructions in the main `README`, these generic programs will likely be in your [PATH](https://en.wikipedia.org/wiki/PATH_%28variable%29).
Their options are displayed via the following command-lines:

```{r functioning_help, eval=FALSE}
model4all_simul --help
model4all_infer --help
model4all_eval --help
```

Currently, these generic programs happens to be scripts coded in the R programming language.
To share information from one step, say `simulate`, to the next, say `infer`, R objects are typically written into compressed files using the [`save`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/save.html) R function, and read using the [`load`](https://stat.ethz.ch/R-manual/R-patched/library/base/html/load.html) R function.



# Simulation step

## Help

The command-line `model4all_simul --help` returns:

```{r simul_help, echo=FALSE}
cat(system(command="../src/model4all_simul --help", intern=TRUE), sep="\n")
```

And for `model4all_simul --version`:

```{r simul_version, echo=FALSE}
cat(system(command="../src/model4all_simul --version", intern=TRUE), sep="\n")
```

## TODO

* give some examples of varuous ways to simulate data



# Inference step

## Help

The command-line `model4all_infer --help` returns:

```{r infer_help, echo=FALSE}
cat(system(command="../src/model4all_infer --help", intern=TRUE), sep="\n")
```


## TODO

* add some details and references in this README about the various possible methods

* for MCMC methods, add option to (try to) detect convergence automatically based on the Monte Carlo standard error (as in [runjags](https://cran.r-project.org/web/packages/runjags/))



# Evaluation step

## Help

The command-line `model4all_eval --help` returns:

```{r eval_help, echo=FALSE}
cat(system(command="../src/model4all_eval --help", intern=TRUE), sep="\n")
```

## Background

### Predictive performances for time series

In the time series context, *one-step-ahead* prediction is of major interest.
It consists in leaving out the last observation and re-fitting the model for predicting the next observation.
The performance of probabilistic prediction can be evaluated using proper scoring rules (Gneiting, 2007; Paul, 2011):

* squared error score (SES): $SES(P,y) = (y-\mu_P)^2$

* logarithmic score (logS): $logS(P,y)=-log(P(Y=y))$

* ranked probability score (RPS): $RPS(P,y)= \sum_{k=0}^{\infty} (P(Y \geq k) - 1(y \geq k))^2$

with $P$ the predictive probability distribution, $\mu_P$ ist first moment, and $y$ the truly observed value.


### Model checking and selection

#### Methods based on the predictive distribution

All measures presented in the following section are based on the predictive distribution.

Let $y^*$ a future value or a value not yet observed in the experiment of the random phenomenon studied through $\boldsymbol{y}$.
By assuming exchangeability for the augmented dataset $\boldsymbol{y*}=\{\boldsymbol{y}, y^*\}$, we have:

\begin{align*}
p(y^* |\boldsymbol{y})& = \frac{p(\boldsymbol{y},y^*)}{p(\boldsymbol{y})}\\
& = \int p(y^*|\theta)p(\theta|\boldsymbol{y}) d\theta,
\end{align*}

with $\theta$ the vector of all parameters. 

#### Cross-validation 

The idea is to divide the sample in two groups: $\boldsymbol{y}=(\boldsymbol{y_t}, \boldsymbol{y_v})$ with $\boldsymbol{y_t}$ used for fitting the model and for estimating the posterior distribution of the parameters, and $\boldsymbol{y_v}$ for evaluating the predictive performance.
Based on the leave one out cross-validation (LOO), assuming that $\boldsymbol{y_t} = \boldsymbol{y_{-i}}$ and $\boldsymbol{y_v} = y_i$, two indices are computed:

1. The conditional predictive ordinate (CPO) \cite{Pettit1990}: $CPO_i = p(y_i^* |\boldsymbol{y_t})$
2. The probability integral transform (PIT) \cite{Dawid1984}: $PIT_i=p(y_i^* \leq y_i | \boldsymbol{y_t})$

For evaluating the predictive performance of the model, we check the empirical distribution of the PIT: a uniform distribution is expected meaning that the predictive distribution is coherent with the data (Gneiting, 2007).
The sum of the log of the CPO values (Carlin, 2008), called the logarithm score, can be also used to compare models in terms of prediction performance (larger values denote a better fitting). 
The Watanabe-Akaike information criterion (WAIC; Watanabe 2010) is an approximation of leave-one-out (LOO) cross-validation.

Indices presented above are implemented in INLA. 

Another view of the logarithm score:
\begin{align*}
logS(P,y) = -log(P(Y=y)),
\end{align*}

with $P$ the predictive probability distribution, and $y$ the truly observed value.
Another measures called the ranked probability score (RPS) not implemented in INLA but used in other approaches seems be a robust alternative:

\begin{align*}
RPS(P, y) = \sum_{k=0}^{\infty} (P(Y \leq k) - 1(y \leq k))^2
\end{align*}


#### Posterior predictive check

The posterior predictive checks have been introduced by Gelman (1996) and are basd on the assumption that $\boldsymbol{y_c} = \boldsymbol{y_f} = \boldsymbol{y}$.
All observations are used for estimating and checking. 
Two quantities are of interest:

1. The posterior predictive distribution $p(y_i^*|\boldsymbol{y}) = \int p(y_{i}^{*}|\theta_i)p(\theta_i|\boldsymbol{y}) d\theta _i$
2. The posterior predicitve p-value desfined as $p(y_i^* \leq y_i | \boldsymbol{y})$

In addition, summary indices can be obtained to globally evaluate the goddness of fit of the model.
The mean square error (MSE) and R squared ($R^2$)


## TODO

* add some details and references in this README about the various possible criteria depending on the inferential methods

* see how to streamline and make more generic this step of the analysis



# How to contribute?

The best way to understand how to contribute is to read this document and have a look at the `R` files for the existing themes already present in the project directory.

If you want to add a new tool at the "inference" step for an already-existing theme, you will need to edit the files named `<theme>_infer.R` and `<theme>_eval.R` in the `theme_<theme>` sub-directory.

If you want to add a theme, you will need to choose a name, and create a sub-directory named `theme_<theme>`.
Then, inside this new sub-directory, you will have to create the following files: `README_theme-<theme>.Rmd`, `<theme>_simul.R`, `<theme>_infer.R` and `<theme>_eval.R`.
Of course, don't hesitate to look up the files from the other themes to save time and avoid doing something too different.

Concretely, you can contribute by:

- reporting [issues](https://github.com/timflutre/model4all/issues);

- proposing [pull requests](https://github.com/timflutre/model4all/pulls).

Note that for any substantial pull request, it's better to create a new [branch](https://www.git-scm.com/book/en/v2/Git-Branching-Branches-in-a-Nutshell).

Finally, don't hesitate to post [issues](https://github.com/timflutre/model4all/issues) on GitHub to let others know so that they can help you.
At this moment, the following persons substantially contributed to the project and can be of some help (contributions following [R guidelines](http://journal.r-project.org/archive/2012-1/RJournal_2012-1_Hornik~et~al.pdf)):

* Timothée Flutre (cre, aut)

* Marie Denis (aut)

* Gabrielle Weinrott (aut)

* ...



# Distributed computations

For some data sets, models, algorithms and/or implementations, the inference step can take quite some time.
In such cases, it is worth using a computer cluster, launching one job per simulated data set.
Here is an example from the "quantgen" theme at the inference step using [rstan](https://cran.r-project.org/web/packages/rstan/) with a computer cluster managing job scheduling via [SGE](https://en.wikipedia.org/wiki/Oracle_Grid_Engine):

```{r infer_qsub_jobs, eval=FALSE}
cd <...>/theme_quantgen/
pkg="rstan"; ls quantgen_simul/* | while read f; do \
  i=$(basename $f | awk '{split($0,a,"\\."); split(a[1],b,"_"); print b[length(b)]}'); \
  echo "model4all_infer --theme quantgen --src <...> --args \"--pkg $pkg\" \
  --task quantgen-AM --simd quantgen_simul --infd quantgen_infer_$pkg --sid ${i}" \
  | qsub -cwd -j y -V -N stdout_infer_quantgen_${pkg}_${i} -q normal.q; \
done
```



# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
