---
title: "Theme \"bpca\""
author: "G. Weinrott"
date: '`r as.character(format(Sys.Date(), format="%d/%m/%Y"))`'
colorlinks: true
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: true
  word_document: default
---

<!--
This R chunk is used to set up important options and load required packages.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")#, fig.width=12, fig.height=8, fig.path=figures/")
```


# Statistical modelling

## References

* Christopher M. Bishop : [Bayesian Principal Component Analysis](http://papers.nips.cc/paper/1549-bayesian-pca.pdf)

* Petr Keil and Jan Smycka : ['bPCA package'](http://www.petrkeil.com/?p=2316)


## Notations

* $N$: number of individuals

* $P$: number of metrics measured during the experiment

* $Y$: data matrix where $\boldsymbol{y}_1$, ..., $\boldsymbol{y}_i$, ..., $\boldsymbol{y}_N$ are the columns, each of length $P$


## Model

We fit a multivariate normal distribution to the $\boldsymbol{y}_i$, so we obtain the following likelihood:

\[
\boldsymbol{y}_i \sim \mathcal{N}_P(\boldsymbol{\mu}, \Sigma)
\]

The priors are the following:

\begin{eqnarray*}
\boldsymbol{\mu} &\sim& \mathcal{N}_P(\boldsymbol{0}_P, \Phi) \\
\Sigma^{-1} &\sim& \mathcal{W}(\Psi, \frac{1}{P})
\end{eqnarray*}

For simplicity, suppose that both $\Phi$ and $\Psi$ are diagonal with variance $P$.
It is certainly possible to treat $\Phi$ and $\Psi$ as hyperparameters, which adds a level to the Bayesian model.



# Simulation

## Help

The command-line `model4all_simul --theme bpca --themhlp` returns:

```{r simul_help, echo=FALSE}
cat(system(command="../src/model4all_simul --theme bpca --themhlp", intern=TRUE), sep="\n")
```


## Case studies

We randomly simulate the data matrix $Y$ as an [Ornstein-Uhlenbeck process](https://en.wikipedia.org/wiki/Ornstein-Uhlenbeck_process).
The user can choose the number of curves to generate, and the number of points of observations for each curve.

Simulate data for the "bpca" theme:

```{r bpca_simul, eval=FALSE}
model4all_simul --theme bpca --simd bpca_simul
```



# Inference

## Help

The command-line `model4all_infer --theme bpca --themhlp` returns:

```{r infer_help, echo=FALSE}
cat(system(command="../src/model4all_infer --theme bpca --themhlp", intern=TRUE), sep="\n")
```


## Case studies

### rstan

```{r infer_rstan, eval=FALSE}
model4all_infer --theme bpca --simd bpca_simul --infd bpca_infer_rstan \
--args "--pkg rstan"
```

### rjags

```{r infer_rjags, eval=FALSE}
model4all_infer --theme bpca --simd bpca_simul --infd bpca_infer_rjags \
--args "--pkg rjags"
```

### TODO

* debug rstan model, not sure where the error comes from...

* rjags seems to use a lot of memory




# Evaluation

## Help

The command-line `model4all_eval --theme bpca --themhlp` returns:

```{r eval_help, echo=FALSE}
cat(system(command="../src/model4all_eval --theme bpca --themhlp", intern=TRUE), sep="\n")
```


## Case studies

Evaluate inference, for instance for `rjags` using the command : 

```{r eval_rjags, eval=FALSE}
pkg="rjags"; model4all_eval --theme bpca --simd bpca_simul \
--infd bpca_infer_${pkg} --evad bpca_eval_${pkg} --args "--pkg ${pkg}"
```

This evaluation step uses Flegal's package to calculate batch means : ['mcmcse'](https://cran.r-project.org/web/packages/mcmcse/mcmcse.pdf) and subsequently produce the Monte Carlo standard error.



# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
